Proof nets for first-order additive linear logic

### 2019-04-23 7 - FSCD 2019 - Final/ALL1.tex
final edit
tweaked spacing to make \relatedversion work --- now everything up to last reference is within the first 17 pages; submitted to server

\author
 {Willem B.\ Heijltjes}
 {University of Bath, United Kingdom}
 {{http://willem.heijltj.es}} %{w.b.heijltjes@bath.ac.uk}
 {}
 {}
\author
 {Dominic J.D.\ Hughes}
 {Logic Group, UC Berkeley, USA}
 {{http://boole.stanford.edu/\textasciitilde dominic}} %{dominich@theory.stanford.edu}
 {}
 {} 
\author
 {Lutz Stra\ss burger}
 {Inria Saclay \&\ LIX/\'Ecole Polytechnique, Palaiseau, France}
 {{http://www.lix.polytechnique.fr/Labo/Lutz.Strassburger}}
 %{lutz.strassburger@inria.fr}
 {}
 {}

@@

%\subjclass{F.4.1: Mathematical logic--Proof theory}
% Theory of computation - Logic - Proof theory
% Theory of computation - Logic - Linear logic
\ccsdesc[500]{Theory of computation~Proof theory}
\ccsdesc[300]{Theory of computation~Linear logic}
\keywords{Linear logic, First-order logic, Proof nets, Herbrand's Theorem}

%\relatedversion{http://hal.inria.fr/hal-01867625/}

@@

%We investigate the \emph{proof-net problem}, finding a canonical representation of proofs, for additive linear logic (\all) extended with first-order quantification (\all{\textnormal 1}). \all\ is the fragment of linear logic \cite{Girard-1987} that describes sums and products, and their canonical morphisms: projections, injections, and diagonals. It is rich, well-behaved, and compact: an ideal microcosm in which to explore semantic and proof-theoretic ideas. Here, we explore two techniques to account for the witness information that is essential to first-order proof: \emph{explicit substitution} and first-order \emph{unification}.

@@

%Additive linear logic $(\all)$ is the logic of product $(\*)$, sum $(\+)$, and their canonical morphisms: projections, injections, and diagonals. It is naturally captured in \defn{proof nets} \cite{Girard-1987,Hughes-vanGlabbeek-2005}, which describe a morphism $A\to B$ by a sequent $\seq{\dual A}B$ plus a \defn{linking}, a relation between the propositional atoms of $\dual A$ and those of $B$. Four examples are illustrated below left.
%
%Additive proof nets are \defn{canonical}: they factor out the permutations of sequent calculus, and correspond 1--1 to morphisms of the free category. \defn{Composition}, of proof nets over $\seq{\dual A}B$ and $\seq{\dual B}C$ to one over $\seq{\dual A}C$, is by the relational composition of their linkings along the dual formulas $B$ and $\dual B$ (as illustrated below right), and captures sequent-calculus cut-elimination.
%
%\[
%\hspace*{9ex}% spacing hack to achieve central horizontal alignment
% Behold my better spacing hack to achieve central horizontal alignment :)
%\hspace*{-\parindent}%
%\hspace*{\fill}%
%\begin{array}{c@{\qquad\qquad}c}
%	\begin{tikzpicture}[net]
%		\formula[y=1]{{\dual a}}
%		\formula[y=0]{a\*a}
%		\Vlink[red]{[-2]1,1}
%		\Vlink[red]{[2]1,3}
%	\end{tikzpicture}
%&
%	\begin{tikzpicture}[net]
%		\formula[y=1]{{\dual a}}
%		\formula[y=0]{a\+b}
%		\Vlink[red]{1,1}
%	\end{tikzpicture}
%\\
%	\text{\small diagonal}
%&	\text{\small injection}	
%\\ \\ \\[-10pt]
%	\begin{tikzpicture}[net]
%		\formula[y=1]{{\dual a}\+{\dual b}}
%		\formula[y=0]{b\*a}
%		\Vlink[red]{1,3}
%		\Vlink[red]{3,1}
%	\end{tikzpicture}
%&
%	\begin{tikzpicture}[net]
%		\formula[y=1]{*{\phantom(}{\dual a}*{\phantom)}
%		 {{\+}}*({\dual b}*{\phantom)}
%		 {{\+}}*{\phantom(}{\dual c}*)}
%		\formula[y=0]{*(a*{\phantom)}
%		{{\*}}*{\phantom(}b*)
%		{{\*}}*{\phantom(}c*{\phantom)}}
%		\Vlink[red]{1,1}
%		\Vlink[red]{3,3}
%		\Vlink[red]{5,5}
%	\end{tikzpicture}
%\\
%	\text{\small symmetry}
%&	\text{\small associativity}
%\end{array}
%\qquad\qquad\qquad
%\begin{array}{c}
%    \vc{\begin{tikzpicture}[net]
%        \formula[y=2.5]{{\dual b}*\+{\dual a}}
%        \formula[y=1.5]{a*\**(b*\+b*)}
%            \Vlink[red]{1,2}
%            \Vlink[red]{2,1}
%        \formula[y=1]{{\dual a}*\+*({\dual b}*\*{\dual b}*)}
%        \formula[y=0]{b*\*b}
%            \Vlink[blue]{[-1]2,[-1]1}
%            \Vlink[blue]{[1]2,[-1]2}
%            \Vlink[blue]{[-1]3,[1]1}
%            \Vlink[blue]{[1]3,[1]2}
%    \end{tikzpicture}}
%\quad\Rightarrow\quad
%    \vc{\begin{tikzpicture}[net]
%        \formula[y=2.5]{{\dual b}*\+{\dual a}}
%        \formula[y=0]{b*\*b}
%            \Vlink[violet]{[-1]1,1}
%            \Vlink[violet]{[1]1,2}
%    \end{tikzpicture}}
%\\ \\[-5pt]
%	\text{\small composition}
%\end{array}
%\hfill
%\]

\paragraph*{Additive proof nets}
%
\all\ proof nets \cite[Section 4.10]{Hughes-vanGlabbeek-2005} represent a morphism $A\to B$ by a sequent $\seq{\dual A}B$ plus a \defn{linking}, a relation between the propositional atoms of $\dual A$ (the dual of $A$) and those of $B$. They are \defn{canonical}: they factor out the permutations of sequent calculus, and correspond 1--1 to morphisms of the free category with binary sums and products. \defn{Composition}, of proof nets over $\seq{\dual A}B$ and $\seq{\dual B}C$ to one over $\seq{\dual A}C$, is by the relational composition of their linkings along the dual formulas $B$ and $\dual B$ and captures sequent-calculus cut-elimination. Below are examples of proof nets and their composition.
%
\[
\hspace*{-\parindent}%
\hspace*{\fill}%
\begin{array}{@{}rc@{\qquad}rc@{}}
	\text{\small diagonal:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\*a}
		\Vlink[red]{[-2]1,1}
		\Vlink[red]{[2]1,3}
	\end{tikzpicture}}
&	\text{\small injection:}	
&   \vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\+b}
		\Vlink[red]{1,1}
	\end{tikzpicture}}
\\ \\
	\text{\small symmetry:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}\+{\dual b}}
		\formula[y=0]{b\*a}
		\Vlink[red]{1,3}
		\Vlink[red]{3,1}
	\end{tikzpicture}}
&	\text{\small associativity:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{*{\phantom(}{\dual a}*{\phantom)}
		 {{\+}}*({\dual b}*{\phantom)}
		 {{\+}}*{\phantom(}{\dual c}*)}
		\formula[y=0]{*(a*{\phantom)}
		{{\*}}*{\phantom(}b*)
		{{\*}}*{\phantom(}c*{\phantom)}}
		\Vlink[red]{1,1}
		\Vlink[red]{3,3}
		\Vlink[red]{5,5}
	\end{tikzpicture}}
\end{array}
\qquad
	\text{\small composition:}
	\quad
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=1.5]{a*\**(b*\+b*)}
            \Vlink[red]{1,2}
            \Vlink[red]{2,1}
        \formula[y=1]{{\dual a}*\+*({\dual b}*\*{\dual b}*)}
        \formula[y=0]{b*\*b}
            \Vlink[blue]{[-1]2,[-1]1}
            \Vlink[blue]{[1]2,[-1]2}
            \Vlink[blue]{[-1]3,[1]1}
            \Vlink[blue]{[1]3,[1]2}
    \end{tikzpicture}}
~~\Rightarrow~~
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=0]{b*\*b}
            \Vlink[violet]{[-1]1,1}
            \Vlink[violet]{[1]1,2}
    \end{tikzpicture}}
\hfill
% Tightening the text and spacing the 
\rule[-55pt]{0pt}{110pt}
\]
%
We extend additive proof nets with first-order quantification. Our central challenge is to incorporate the essential content of first-order proof, the \defn{witness assignment} to existential quantifiers. Commonly, as in the sequent calculus rule below left, a witness to $\exists x.B$ is given by an immediate substitution $B[t/x]$. To assign different witnesses in different branches of a proof, the subformula is duplicated first, giving $B[s/x]$ and $B[t/x]$, as below right.
%
\[
\hspace*{\fill}
\hspace*{-\parindent}
	\vc{\infer[\Qrr?t]{\seq A{\exists x.B}}{\seq A{B[t/x]}}}
\hfill
\vc{
	\infer[\Qrr*]{\Seq{\exists x.\dual P(x)}{P(s)\*P(t)}}{
	 \infer[\Qrr?s]{\Seq{\exists x.\dual P(x)}{P(s)}}{\infer{\Seq{\dual P(s)}{P(s)}}{}}
	 &&
	 \infer[\Qrr?t]{\Seq{\exists x.\dual P(x)}{P(t)}}{\infer{\Seq{\dual P(t)}{P(t)}}{}}
   }}
\hfill
\]
%
This is incompatible with a \defn{sequent\,+\,links} proof net design, where the conclusion sequent remains intact, and a subformula $B$ cannot be the subject of substitution or duplication. Instead, we propose two alternative treatments of witnessing terms, embodied in two notions of proof net: \defn{witness nets} and \defn{unification nets}. Our solutions are based on the second author's recent \defn{unification nets} for first-order multiplicative linear logic \cite{Hughes-2018}. Their main feature is to omit existential witnesses altogether, and reconstruct them by unification.
%

@@

\defn{Unification nets} omit any witness information, as illustrated below right. In addition to canonicity, they embody a notion of \defn{generality}: where more than one witness could be assigned, unification nets do not require a definite choice, while witness nets do. Composition is direct, by relational composition. We compare further properties in Figure~\ref{fig:results} in the conclusion, where we also discuss related work and Lambek's notion of \defn{generality} \cite{Lambek-1968-1972}.
%\cite{Lambek-1968, Lambek-1969, Lambek-1972}.
%
\[
\hspace*{\fill}
\hspace*{-\parindent}
	\text{\small a witness net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red,l,label={$\scriptstyle[s/x]~~$}]{[-3]4,1}
		\Vlink[red,r,label={$~\scriptstyle[t/x]$}]{[1]4,6}
	\end{tikzpicture}}
\hfill
	\text{\small a unification net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red]{[-3]4,1}
		\Vlink[red]{[1]4,6}
	\end{tikzpicture}}
\hfill
\]
%
\paragraph*{Background}
%
Additive linear logic is combinatorially rich, yet well-behaved and tractable: proof search \cite{Galmiche-Marion-1995} and proof net correctness \cite{Heijltjes-Hughes-2015} for a net over $\seq AB$ are linear in $|A|\times|B|$ (with $|A|$ the size of the syntax tree of $A$). Proof nets remain canonical and equally tractable when extended with the two units \cite{Heijltjes-2011,Heijltjes-Hughes-2015}, and the first-order case is merely NP-complete \cite{Heijltjes-Hughes-2015}.

%At the same time, the ubiquity of its main operations, product and sum, mean the ideas and results it garners are widely applicable. It serves as a microcosmos for important ideas and observations: it fostered the \emph{connections method} of proof search \cite{Galmiche-2000}; it demonstrated the \emph{Blass problem} \cite{Abramsky-2003} of game semantics, that sequential strategies do not in general have associative composition; and it is at the root of the study of fixed points in linear logic \cite{Santocanale-2002a}. The logic describes two-way communication~\cite{Cockett-Santocanale-2009}, and is a core part of \emph{session types}~\cite{Honda-1993}.

{\all} is of course part of {\mall} (multiplicative-additive linear logic), and its lessons are clearly visible in the second author's canonical {\mall} proof nets~\cite{Hughes-vanGlabbeek-2005}, as well as the first and second author's locally canonical \emph{conflict nets}~\cite{Hughes-Heijltjes-2016}. Its proof nets also appear in the third author's study of the \emph{medial rule} for classical logic~\cite{str:RTA07}, and as the \emph{skew fibrations} in the second author's \emph{combinatorial proofs} for classical logic~\cite{Hughes-2006-proofs}. To prepare the ground for cut elimination in first-order \emph{combinatorial proofs} is a further motivation for the present work.

@@

We will distinguish \defn{strict coalescence} ($\scoal$) for witness nets and \defn{unifying coalescence} ($\ucoal$) for unification nets. We first give an example of the former (abbreviating $P(x,y)$ to $Pxy$).
%(writing $Pxy$ for $P(x,y)$ to conserve space). 
In the initial witness net, below left, each link corresponds to a sequent calculus axiom between both linked subformulas, with the substitutions applied.
%
\newcommand\xs{xs} %{(\kern-1ptx\kern-1pt,\kern-2pts\kern-1pt)}
\newcommand\xt{xt} %{(\kern-1ptx\kern-1pt,\kern-2ptt\kern-1pt)}
\newcommand\xy{xy} %(\kern-1ptx\kern-1pt,\kern-2pty\kern-1pt)}
\newcommand\zs{zs}
\newcommand\zt{zt}
\[
\hspace*{-\parindent}
\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z,s/y]~~$}]{[-3]3,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \quad\scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \scoal &
\hfill
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z]$}]{[2]2,4}
	\end{tikzpicture}}
\hfill
\scoal\cdots
\\ \\
\text{\scriptsize
	\color{white}
	 \infer{\vphantom{\dual P}}{
	  \color{red}\infer{\seq{\dual P\xs}{P\xs}}{}
       &&
	  \color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}
	 }}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{white}\infer{\vphantom{\dual P}}{\color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}}}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{blue}\infer[\!\Qrr?t]{\seq{\exists y.\dual P\xy}{P\xt}}{\color{black}\infer{\seq{\dual P\xt}{P\xt}}{}}}
\end{array}
\]
The first two steps (above middle and right) move both links from the subformula $\dual Pxy$ to $\exists y.\dual Pxy$, removing the substitutions $[s/y]$ and $[t/y]$ on $y$, and corresponding to sequent rules $\qrr?s$ and $\qrr?t$. Observe that we maintain the domain of the substitutions on a link as the variables of those existential quantifiers that either linked subformula is (strictly) in scope of.

>>>>

\author
 {Willem B.\ Heijltjes}
 {University of Bath, United Kingdom}
 {{http://willem.heijltj.es}} %{w.b.heijltjes@bath.ac.uk}
 {}
 {}
\author
 {Dominic J.\ D.\ Hughes}
 {Logic Group, UC Berkeley, USA}
 {{http://boole.stanford.edu/\textasciitilde dominic}} %{dominich@theory.stanford.edu}
 {}
 {} 
\author
 {Lutz Stra\ss burger}
 {Inria Saclay \&\ LIX/\'Ecole Polytechnique, Palaiseau, France}
 {{http://www.lix.polytechnique.fr/Labo/Lutz.Strassburger}}
 %{lutz.strassburger@inria.fr}
 {}
 {}

@@

\ccsdesc[500]{Theory of computation~Proof theory}
\ccsdesc[300]{Theory of computation~Linear logic}
\keywords{Linear logic, First-order logic, Proof nets, Herbrand's Theorem}

\relatedversion{A full version \cite{Heijltjes-Hughes-Strassburger-2018} is available at \url{http://hal.inria.fr/hal-01867625/}}

@@

@@

\paragraph*{Additive proof nets}
%
\all\ proof nets \cite[Section 4.10]{Hughes-vanGlabbeek-2005} represent a morphism $A\to B$ by a sequent $\seq{\dual A}B$ plus a \defn{linking}, a relation between the propositional atoms of $\dual A$ (the dual of $A$) and those of $B$. They are \defn{canonical}: they factor out the permutations of sequent calculus, and correspond 1--1 to morphisms of the free category with binary sums and products. \defn{Composition}, of proof nets over $\seq{\dual A}B$ and $\seq{\dual B}C$ to one over $\seq{\dual A}C$, is by the relational composition of their linkings along the dual formulas $B$ and $\dual B$ and captures sequent-calculus cut-elimination. Below are examples of proof nets and their composition.
%
\begin{center}\vspace{-.5ex}\(
\begin{array}{@{}rc@{\qquad}rc@{}}
	\text{\small diagonal:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\*a}
		\Vlink[red]{[-2]1,1}
		\Vlink[red]{[2]1,3}
	\end{tikzpicture}}
&	\text{\small injection:}	
&   \vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}}
		\formula[y=0]{a\+b}
		\Vlink[red]{1,1}
	\end{tikzpicture}}
\\ \\
	\text{\small symmetry:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{{\dual a}\+{\dual b}}
		\formula[y=0]{b\*a}
		\Vlink[red]{1,3}
		\Vlink[red]{3,1}
	\end{tikzpicture}}
&	\text{\small associativity:}
&	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{*{\phantom(}{\dual a}*{\phantom)}
		 {{\+}}*({\dual b}*{\phantom)}
		 {{\+}}*{\phantom(}{\dual c}*)}
		\formula[y=0]{*(a*{\phantom)}
		{{\*}}*{\phantom(}b*)
		{{\*}}*{\phantom(}c*{\phantom)}}
		\Vlink[red]{1,1}
		\Vlink[red]{3,3}
		\Vlink[red]{5,5}
	\end{tikzpicture}}
\end{array}
\qquad
	\text{\small composition:}
	\quad
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=1.5]{a*\**(b*\+b*)}
            \Vlink[red]{1,2}
            \Vlink[red]{2,1}
        \formula[y=1]{{\dual a}*\+*({\dual b}*\*{\dual b}*)}
        \formula[y=0]{b*\*b}
            \Vlink[blue]{[-1]2,[-1]1}
            \Vlink[blue]{[1]2,[-1]2}
            \Vlink[blue]{[-1]3,[1]1}
            \Vlink[blue]{[1]3,[1]2}
    \end{tikzpicture}}
~~\Rightarrow~~
    \vc{\begin{tikzpicture}[net]
        \formula[y=2.5]{{\dual b}*\+{\dual a}}
        \formula[y=0]{b*\*b}
            \Vlink[violet]{[-1]1,1}
            \Vlink[violet]{[1]1,2}
    \end{tikzpicture}}
\rule[-55pt]{0pt}{110pt}
\)\vspace{-.5ex}\end{center}
%
We extend additive proof nets with first-order quantification. Our central challenge is to incorporate the essential content of first-order proof, the \defn{witness assignment} to existential quantifiers. Commonly, as in the sequent calculus rule below left, a witness to $\exists x.B$ is given by an immediate substitution $B[t/x]$. To assign different witnesses in different branches of a proof, the subformula is duplicated first, giving $B[s/x]$ and $B[t/x]$, as below right.
%
\begin{center}\(
%\hspace*{\fill}
%\hspace*{-\parindent}
	\vc{\infer[\Qrr?t]{\seq A{\exists x.B}}{\seq A{B[t/x]}}}
\hspace{10ex}
\vc{
	\infer[\Qrr*]{\Seq{\exists x.\dual P(x)}{P(s)\*P(t)}}{
	 \infer[\Qrr?s]{\Seq{\exists x.\dual P(x)}{P(s)}}{\infer{\Seq{\dual P(s)}{P(s)}}{}}
	 &&
	 \infer[\Qrr?t]{\Seq{\exists x.\dual P(x)}{P(t)}}{\infer{\Seq{\dual P(t)}{P(t)}}{}}
   }}
%\hfill
\)\end{center}
%
This is incompatible with a \defn{sequent\,+\,links} proof net design, where the conclusion sequent remains intact, and a subformula $B$ cannot be the subject of substitution or duplication. Instead, we propose two alternative treatments of witnessing terms, embodied in two notions of proof net: \defn{witness nets} and \defn{unification nets}. Our solutions are based on the second author's recent \defn{unification nets} for first-order multiplicative linear logic \cite{Hughes-2018}. Their main feature is to omit existential witnesses altogether, and reconstruct them by unification.
%

@@

\defn{Unification nets} omit any witness information, as illustrated below right. In addition to canonicity, they embody a notion of \defn{generality}: where more than one witness could be assigned, unification nets do not require a definite choice, while witness nets do. Composition is direct, by relational composition. We compare further properties in Figure~\ref{fig:results} in the conclusion, where we also discuss related work and Lambek's notion of \defn{generality} \cite{Lambek-1968-1972}.
%
\begin{center}\(
	\text{\small a witness net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red,l,label={$\scriptstyle[s/x]~~$}]{[-3]4,1}
		\Vlink[red,r,label={$~\scriptstyle[t/x]$}]{[1]4,6}
	\end{tikzpicture}}
\hspace{11ex}
	\text{\small a unification net:}
	\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1.2]{\exists x.{\dual P}(x)}
		\formula[y=0]{P(s)\*P(t)}
		\Vlink[red]{[-3]4,1}
		\Vlink[red]{[1]4,6}
	\end{tikzpicture}}
\)\vspace{-1ex}\end{center}
%
\paragraph*{Background}
%
Additive linear logic is combinatorially rich, yet well-behaved and tractable: proof search \cite{Galmiche-Marion-1995} and proof net correctness \cite{Heijltjes-Hughes-2015} for a net over $\seq AB$ are linear in $|A|\times|B|$ (with $|A|$ the size of the syntax tree of $A$). Proof nets remain canonical and equally tractable when extended with the two units \cite{Heijltjes-2011,Heijltjes-Hughes-2015}, and the first-order case is merely NP-complete \cite{Heijltjes-Hughes-2015}.

{\all} is of course part of {\mall} (multiplicative-additive linear logic), and its lessons are clearly visible in the second author's canonical {\mall} proof nets~\cite{Hughes-vanGlabbeek-2005}, as well as the first and second author's locally canonical \emph{conflict nets}~\cite{Hughes-Heijltjes-2016}. Its proof nets also appear in the third author's study of the \emph{medial rule} for classical logic~\cite{str:RTA07}, and as the \emph{skew fibrations} in the second author's \emph{combinatorial proofs} for classical logic~\cite{Hughes-2006-proofs}. To prepare the ground for cut elimination in first-order combinatorial proofs~\cite{Hughes-2019-fopws} is further motivation for the present work.

@@

We will distinguish \defn{strict coalescence} ($\scoal$) for witness nets and \defn{unifying coalescence} ($\ucoal$) for unification nets. We first give an example of the former (abbreviating $P(x,y)$ to $Pxy$). In the initial witness net, below left, each link corresponds to a sequent calculus axiom between both linked subformulas, with the substitutions applied.
%
\newcommand\xs{xs}%
\newcommand\xt{xt}%
\newcommand\xy{xy}%
\newcommand\zs{zs}%
\newcommand\zt{zt}%
\begin{center}\vspace{1pt}\(
\hspace*{-\parindent}
\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z,s/y]~~$}]{[-3]3,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \quad\scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,t/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \scoal &
\hfill
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z]$}]{[2]2,4}
	\end{tikzpicture}}
\hfill
\scoal\cdots
\\[6ex]
\text{\scriptsize
	\color{white}
	 \infer{\vphantom{\dual P}}{
	  \color{red}\infer{\seq{\dual P\xs}{P\xs}}{}
       &&
	  \color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}
	 }}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{white}\infer{\vphantom{\dual P}}{\color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}}}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{blue}\infer[\!\Qrr?t]{\seq{\exists y.\dual P\xy}{P\xt}}{\color{black}\infer{\seq{\dual P\xt}{P\xt}}{}}}
\end{array}
\)\vspace{-3pt}\end{center}
The first two steps (above middle and right) move both links from the subformula $\dual Pxy$ to $\exists y.\dual Pxy$, removing the substitutions $[s/y]$ and $[t/y]$ on $y$, and corresponding to sequent rules $\qrr?s$ and $\qrr?t$. Observe that we maintain the domain of the substitutions on a link as the variables of those existential quantifiers that either linked subformula is (strictly) in scope of.

The next step, from above right to below left, combines both links, and corresponds to an additive conjunction rule. We require that both substitutions agree (their domains are the same by the above observation); hence this step could not have been performed before the previous two, corresponding to the non-permutability of the generated inference rules.
\begin{center}\(
\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}}
\cdots\scoal\quad
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[violet,r,label={$~~\scriptstyle[x/z]$}]{2,3}
	\end{tikzpicture}}
	\hfill\hfill
& \scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[violet]{2,1}
	\end{tikzpicture}}
& \scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[violet]{1,1}
	\end{tikzpicture}}
\\[6ex]
\text{\scriptsize
	\color{white}\infer{\dual P}{\infer{\dual P}{
	\color{violet}
	 \infer[\Qrr*]{\seq{\exists y.\dual P\xy}{P\xs\*P\xt}}{
	  \color{black}\infer{\seq{\exists y.\dual P\xy}{P\xs}}{\infer{\seq{\dual P\xs}{P\xs}}{}}
	  &
	  \color{black}\infer{\seq{\exists y.\dual P\xy}{P\xt}}{\infer{\seq{\dual P\xt}{P\xt}}{}}
	 }
	}}}
& \hspace*{28pt} &
\text{\scriptsize
	\color{white}\infer{\dual P}{
	\color{violet}
 	\infer[\Qrr?x]{\seq{\exists y.\dual P\xy}{\exists z.P\zs\*P\zt}}{
	 \color{black}
	 \infer{\seq{\exists y.\dual P\xy}{P\xs\*P\xt}}{
	  \infer{\seq{\exists y.\dual P\xy}{P\xs}}{\infer{\seq{\dual P\xs}{P\xs}}{}}
	  &
	  \infer{\seq{\exists y.\dual P\xy}{P\xt}}{\infer{\seq{\dual P\xt}{P\xt}}{}}
	}}
	}}
& \hspace*{28pt} &
\text{\scriptsize
	\color{violet}
	\infer[\Qrr!]{\seq{\forall x.\exists y.\dual P\xy}{\exists z.P\zs\*P\zt}}{
	 \color{black}
 	 \infer{\seq{\exists y.\dual P\xy}{\exists z.P\zs\*P\zt}}{
	  \infer{\seq{\exists y.\dual P\xy}{P\xs\*P\xt}}{
	   \infer{\seq{\exists y.\dual P\xy}{P\xs}}{\infer{\seq{\dual P\xs}{P\xs}}{}}
	   &
	   \infer{\seq{\exists y.\dual P\xy}{P\xt}}{\infer{\seq{\dual P\xt}{P\xt}}{}}
	}}}}
\end{array}
\)\end{center}
The final steps introduce an inference $\exists R,x$ and one $\forall R$. For the latter, we require that the \emph{eigenvariable} $x$ of the universal quantification does not occur in the range of the substitution of the link, as in $[x/z]$ in the net above left---hence the two steps could not be interchanged. It corresponds to the \emph{eigenvariable condition} on the $\forall R$ rule that $x$ is not free in the context.

Unifying coalescence, for unification nets, is similar to strict coalescence; two differences allow it to reconstruct witnesses by unification, which we illustrate. To initialize coalescence, the links in the net below left are given, as a substitution map, the \defn{most general unifier} of the two propositions connected by the link. Both links now correspond to sequent axioms.
\begin{center}\vspace{0pt}\(
\hspace*{-\parindent}
\hspace*{\fill}
\begin{array}{@{}ccccc@{}}
    \vc{\begin{tikzpicture}[net]
    	\formula[y=1]{{\dual P}*{xx}}
    	\formula[y=0]{P*{ys}\*P*{tz}}
    	\Vlink[red]{[-2]1,1}
    	\Vlink[blue]{[2]1,3}
    \end{tikzpicture}}
&\ucoal& 
    \vc{\begin{tikzpicture}[net]
    	\formula[y=1]{{\dual P}*{xx}}
    	\formula[y=0]{P*{ys}\*P*{tz}}
    	\Vlink[red,l,label={$\scriptstyle[s/y,s/x]$~~}]{[-2]1,1}
    	\Vlink[blue,r,label={~~$\scriptstyle[t/z,t/x]$}]{[2]1,3}
    \end{tikzpicture}}
&\ucoal& 
    \vc{\begin{tikzpicture}[net]
    	\formula[y=1]{{\dual P}*{xx}}
    	\formula[y=0]{P*{ys}\*P*{tz}}
    	\Vlink[violet,r,label={~~$\scriptstyle[u/z,u/y,u/x]$}]{1,2}
    \end{tikzpicture}}
\\[6ex]
&\qquad\qquad&
\text{\scriptsize
	\color{white}
	 \infer{\vphantom{\dual P}}{
	  \color{red}\infer{\seq{\dual Pss}{Pss}}{}
       &&
	  \color{blue}\infer{\seq{\dual Ptt}{Ptt}}{}
	 }}
&\qquad\qquad&
\text{\scriptsize
	\color{violet}
	 \infer[\!\Qrr*]{\seq{\dual Puu}{Puu\*Puu}}{
	  \color{black}\infer{\seq{\dual Puu}{Puu}}{}
       &&
	  \color{black}\infer{\seq{\dual Puu}{Puu}}{}
	 }}
\end{array}
\hfill
\)\vspace{-2pt}\end{center}
The second difference is the coalescence step for additive conjunction, above right. Where strict coalescence requires both links to carry identical substitution maps, here we require both maps to be \emph{unifyable}, in the sense that both must have a common, more special (less general) substitution map. This is then applied to the new link. In the example, the terms $s$ and $t$ are unified to $u$, i.e.\ $u$ is a most general term that specializes both $s$ and $t$. Note that in the sequentialization, both subproofs also need to be specialized, from $s$ and $t$ to $u$.
